A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-09-06 doc/jcomp2012/jcomp2012.tex
Fixed preconditioning matrix.
Added technical editors comments.

    We present a novel synthetic-acceleration based Monte Carlo method for
    solving the equilibrium thermal radiation diffusion equation in three
    spatial dimensions. The algorithm performance is compared against
    traditional solution techniques using a Marshak benchmark problem and a
    more complex multiple material problem. Our results show that our Monte
    Carlo method is an effective solver for sparse matrix systems. For
    solutions converged to the same tolerance, it performs competitively with
    deterministic methods including preconditioned conjugate gradient. We also
    discuss various aspects of preconditioning the method and its general
    applicability to broader classes of problems.

@@

In previous work~\cite{evans_2003}, we developed a residual Monte Carlo method
that was an application of Halton's Sequential Monte Carlo
method~\cite{halton_1962,halton_1994} to the one-dimensional equilibrium
thermal radiation diffusion equation.  At that time, Halton's method was not
widely known in the general computational physics community.  While extending
our method to three spatial dimensions, we have realized that the Sequential
Monte Carlo method is actually a variant of iterative refinement cast as a
residual method. We have developed a new method in which Monte Carlo is used
to accelerate a fixed-point (Richardson) iteration. This algorithm surpasses
our previous one dimensional method and allows for extension to more general
solution techniques.

The purpose of this study is to demonstrate an efficient Monte Carlo solution
method for three dimensional (3D), time-dependent, discrete systems. In this
work, our previous investigations are extended in the following manner:
\begin{itemize}
\item we develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems;
\item we apply this method to the one-temperature (1T) thermal radiation
  diffusion equation in three dimensions;
\item we compare numerical and performance results of the Monte Carlo
  method against traditional deterministic solution methods.
\end{itemize}

In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for solving
matrix systems, and we discuss our new synthetic-acceleration scheme and
connect it with previous work by Halton~\cite{halton_1994} and
others~\cite{evans_2003}.  The 1T radiation diffusion equation and its
discrete form will be derived in \S~\ref{sec:discr-form-radi}.  We discuss
using Monte Carlo Synthetic-Accleration to solve the discrete radiation
diffusion equation in \S~\ref{sec:solut-radi-diff}.  Results from two 3D test
problems are given in \S~\ref{sec:results}.  We finish with conclusions and
ideas for future work in \S~\ref{sec:conclusions}.

@@

The idea of using Monte Carlo methods (random walks) to invert linear
systems is not new.  The earliest referenceable work dates to a 1950
paper by Forsythe and Leibler \cite{forsythe}.  Their paper credits
the idea to unpublished work by J.v. Neumann and S.M. Ulam dating back
to the 1940's.  The basic principles of Monte Carlo matrix inversion
were further elucidated in Hammersley and Handscomb's 1964 text
\cite{hammersley_1964}.  These early methods are distinguished by very
slow, statistically noisy convergence properties; thus, they have not
made any significant impact in the physics community.

In order to address the convergence issues plaguing Monte Carlo
solvers, Halton \cite{halton_1962,halton_1994} proposed a Sequential
Monte Carlo method.  This algorithm demonstrated dramatically improved
convergence over regular Monte Carlo.  Nonetheless, Halton's method
has not gained widespread use in computational physics
applications. In the next sections, we will briefly describe the Monte
Carlo methods that can be used to solve linear systems and expand them
to present our new synthetic acceleration method.

@@

To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius ($\rho$) of the iteration matrix ($\vH$) is
less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
  Series},
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation} 
Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
a series,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
With this knowledge in hand, we can write an iterative method that
solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb,\vx^0 \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the Monte
Carlo methods that are described in \S\S~\ref{sec:direct-method} through
\ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

@@

Now we consider a Monte Carlo method that can be used to estimate the
solution to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First,
rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
component of $\vx$,
\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\ &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}
The terms in the Neumann series can be interpreted as a series of
transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk
with $k$ events that initiates in state $i$,
\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\ &=
    \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}
Here, the weight on the $m^\text{th}$ step is denoted $W_m$, and each
random walk starts with unit weight. At every step, contributions to
each component of $X$ are generated only in the starting state of the
history, $i_0$, by the source term in the current state,
$b_{i_m}$. Therefore, for a given random walk permutation, each state
must be used as a starting value in order to calculate its component
of the expectation value (i.e. a state of size $N$ will have $N$
histories per random walk permutation). When there is a transition
between states, for example $i\rightarrow j$, the weight is multiplied
by the transition factor
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
  \label{eq:weight}
\end{equation}
where $p_{ij}$ denotes the probability of transitioning from state $i$
to $j$. Then, the expected value of $X$ is
\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\ &=
    \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}
where $\nu$ denotes a particular random walk permutation.  Therefore,
the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased estimator
of the components of $\vx$ provided $\rho(\vH) < 1$.

@@

%%---------------------------------------------------------------------------%%
\subsection{Monte Carlo Synthetic-Acceleration}
\label{sec:iter-refin-monte}

The methods presented in \S~\ref{sec:background} are characterized by slow
convergence rates bound by the Central Limit Theorem.
Halton~\cite{halton_1962,halton_1994} proposed a staged residual scheme called
Sequential Monte Carlo to speed up the convergence of these methods.  A
variant of this scheme has been successfully applied to the 1D non-linear
thermal radiation diffusion equation in Ref.~\cite{evans_2003}.

Concisely, the Sequential Monte Carlo method solves Eq.~(\ref{eq:Ax=b}) using
the adjoint solution technique described in \S~\ref{sec:adjoint-method}.  The
following iteration scheme is applied:
\begin{subequations}
  \begin{gather}
    \vr^{l} = \vb - \vA\vx^{l}\:, \\ \vA\delta\vx^{l+1} =
    \vr^{l}\:, \label{eq:residual_solve}\\ \vx^{l+1} = \vx^{l} +
    \delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
In this scheme, the Monte Carlo adjoint method is used to estimate the
solution to Eq.~(\ref{eq:residual_solve}), and the residual is iterated to
convergence.  This iteration sequence is closely related to iterative
refinement; the exception being that there is no update to $x^{l+1}$ between
residual iterations.

We propose a modification of this scheme that uses Monte Carlo as a
synthetic-acceleration for the fixed-point iteration given by
Eq.~(\ref{eq:richardson}). We begin by subtracting Eq.~(\ref{eq:richardson})
from Eq:~(\ref{eq:point_iteration}):
\begin{equation}
  \delta \vx^{l+1} = (\vI - \vA)\delta \vx^{l}\:,
  \label{eq:synthetic_error}
\end{equation}
where
\begin{equation}
  \delta \vx^{l} = \vx - \vx^{l}
\end{equation}
is defined as the error at iteration $l$. We then subtract $(\vI-\vA)
\delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}), giving a linear
system to solve for the error:
\begin{equation}
  \begin{split}
    \vA \delta \vx^{l+1} &= (\vI - \vA)(\vx^{l+1} - \vx^{l}) \\ &=
    \vr^{l+1}\:.
  \end{split}
  \label{eq:synthetic_error2}
\end{equation}

The following scheme will then converge in one iteration:
\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\ \vA \delta \vx^{l+1} =
    \vr^{l+1}\;,
    \label{eq:synthetic-residual-solve}\\ 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
To accelerate the fixed-point method and avoid inverting $\vA$ directly in
Eq.~(\ref{eq:synthetic-residual-solve}), we can instead use the above ideas to
create a new iterative scheme that applies the Monte Carlo method to estimate
the solution error. With this in hand, the Fixed-Point Monte Carlo
Synthetic-Acceleration (MCSA) method can be defined as follows:
\begin{subequations}
  \label{eq:MCSA-iteration}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\  \vr^{l+1/2} = \vb -
    \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\     
     \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
     \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
\end{subequations}
The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that the
Monte Carlo solution only approximately inverts this operator.  Thus, we have
defined a scheme in which the initial estimate of $\vx$ in each iteration is
updated using a single fixed-point iteration\footnote{Although, a subspace
  method could be used instead of a stationary method here.}.  The residual is
calculated and used as a source to estimate the error, $\delta\vx^{l+1/2}$, by
solving Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the solution
vector, $\vx^{l+1}$.  The entire sequence is iterated to convergence based on
the following stopping criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

The MCSA method defined in Eq.~(\ref{eq:MCSA-iteration}) uses the adjoint
method to estimate the error in Eq.~(\ref{eq:MCSA-residual_solve}) instead of
the direct method outlined in \S~\ref{sec:direct-method}. To demonstrate the
effectiveness of the adjoint method over the direct method within the context
of MCSA, we choose the two-dimensional time-dependent Poisson equation as a
simple model problem:
\begin{equation}
  \frac{\partial \vu}{\partial t} = \nabla^2 \vu\:.
  \label{eq:poisson_equation}
\end{equation}
For all comparisons, a single time step is computed with backwards Euler time
integration. The Laplacian is differenced on a square Cartesian grid with a
second-order five point stencil
\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]\:,
  \label{eq:five_point_stencil}
\end{equation}
and a fourth-order nine point stencil 
\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
  \label{eq:nine_point_stencil}
\end{multline}
both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
a single time step solution, we then have the following sparse linear system
to be solved with the MCSA method:
\begin{equation}
  \vA \vu^{n+1} = \vu^n\:.
  \label{eq:poisson_eq_lin_sys}
\end{equation}
Both the stencils will be used to vary the size and density of the sparse
linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).

A timing study and convergence study is used to demonstrate the effectiveness
of the adjoint method as compared to the direct method. To assess both the CPU
time and number of iterations required to converge to a solution, a problem of
constant $\Delta$ was used with varying values of grid size, fixing the
spectral radius of the system at a constant value for each variation. Both the
five point and nine point stencils were used with both the direct and adjoint
solvers. For each case, 50 random walk permutations were computed per MCSA
iteration with a weight cutoff of \sn{1}{-4} for the stochastic linear solver
and a convergence tolerance of \sn{1}{-8} for the MCSA iterative solver. All
computations presented in this section were completed on a 2.66 GHz Intel Core
i7 machine with 8 GB 1067 MHz DDR3 memory. Fig.~\ref{fig:poisson_cpu_time}
gives the CPU time needed for each case to converge in seconds and
Fig.~\ref{fig:poisson_iterations} gives the number of iterations needed for
each case to converge relative to the problem size.
\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five point and nine point stencils. A significant
    CPU time speedup is noted with the adjoint method due to the
    reduced number of random walk events.}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.pdf}
  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N
    \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five point and nine point stencils. Both methods
    give similar iteration behavior.}
  \label{fig:poisson_iterations}
\end{figure}

We see clearly in Fig.~\ref{fig:poisson_cpu_time} that the using the adjoint
solver with MCSA results in several orders of magnitude speedup over the
direct solver while the number of iterations required to converge is of a
similar scale. We expect this for several reasons. First, with an equivalent
number of histories specified for both solvers and a system of size $N \times
N$, the direct solver will compute $N \times N$ random walks per permutation
while the adjoint solver will only compute one. This is necessary in the
direct method to ensure a contribution from each state as the random walk
sequence will only contribute to the starting state. For the adjoint method,
because the random walk sequence contributes to the state in which it
currently resides, fewer histories are neccessary to compute contributions
from all the nonzero parts of the source. Initially, this may not seem like a
fair comparison, however, we see from Fig.~\ref{fig:poisson_iterations} that
the number of iterations required to converge is approximately the same and
therefore the extra computations performed by the direct method do not improve
the error estimation. Therefore, although the total number of random walks per
iteration is in fact larger for the direct method, the solution computed at
each iteration is no better than that computed by the adjoint method.

>>>>

    We present a novel synthetic-acceleration-based Monte Carlo method for
    solving the equilibrium thermal radiation diffusion equation in three
    spatial dimensions. The algorithm performance is compared against
    traditional solution techniques using a Marshak benchmark problem and a
    more complex multiple material problem. Our results show that our Monte
    Carlo method is an effective solver for sparse matrix systems. For
    solutions converged to the same tolerance, it performs competitively with
    deterministic methods including preconditioned conjugate gradient. We also
    discuss various aspects of preconditioning the method and its general
    applicability to broader classes of problems.

@@

In previous work~\cite{evans_2003}, we developed a residual Monte Carlo method
that was an application of Halton's Sequential Monte Carlo
method~\cite{halton_1962,halton_1994} to the one-dimensional (1D) equilibrium
thermal radiation diffusion equation.  At that time, Halton's method was not
widely known in the general computational physics community.  While extending
our method to three spatial dimensions, we realized that the Sequential Monte
Carlo method is actually a variant of iterative refinement cast as a residual
method. We have developed a new method in which Monte Carlo is used to
accelerate a fixed-point (Richardson) iteration. This algorithm surpasses our
previous 1D method and allows for extension to more general solution
techniques.

The purpose of this study is to demonstrate an efficient Monte Carlo solution
method for three-dimensional (3D), time-dependent, discrete systems. In this
work, our previous investigations are extended in the following manner:
\begin{itemize}
\item we develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems;
\item we apply this method to the one-temperature (1T) thermal radiation
  diffusion equation in three dimensions;
\item we compare numerical and performance results of the Monte Carlo
  method against traditional deterministic solution methods.
\end{itemize}

In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for solving
matrix systems, and we discuss our new synthetic-acceleration scheme and
connect it with previous work by Halton~\cite{halton_1994} and
others~\cite{evans_2003}.  The 1T radiation diffusion equation and its
discrete form will be derived in \S~\ref{sec:discr-form-radi}.  We discuss
using Monte Carlo Synthetic-Acceleration to solve the discrete radiation
diffusion equation in \S~\ref{sec:solut-radi-diff}.  Results from two 3D test
problems are given in \S~\ref{sec:results}.  We finish with conclusions and
ideas for future work in \S~\ref{sec:conclusions}.

@@

The idea of using Monte Carlo methods (random walks) to invert linear systems
is not new.  The earliest referenced work dates to a 1950 paper by Forsythe
and Leibler \cite{forsythe}.  Their paper credits the idea to unpublished work
by J.v. Neumann and S.M. Ulam dating back to the 1940's.  The basic principles
of Monte Carlo matrix inversion were further elucidated in Hammersley and
Handscomb's 1964 text \cite{hammersley_1964}.  These early methods are
distinguished by very slow, statistically noisy convergence properties; thus,
they have not made any significant impact in the physics community.

In order to address the convergence issues plaguing Monte Carlo
solvers, Halton \cite{halton_1962,halton_1994} proposed a Sequential
Monte Carlo method.  This algorithm demonstrated dramatically improved
convergence over regular Monte Carlo.  Nonetheless, Halton's method
has not gained widespread use in computational physics
applications. In the next sections, we will briefly describe the Monte
Carlo methods that can be used to solve linear systems and expand them
to present our new synthetic-acceleration method.

@@

To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written as
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius ($\rho$) of the iteration matrix ($\vH$) is
less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
  Series}:
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation} 
Thus, when $\rho(\vH) < 1$, we can recast the solution vector $\vx$ as
a series:
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
With this knowledge in hand, we can write an iterative method that
solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration):
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb,\vx^0 \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the Monte
Carlo methods that are described in \S\S~\ref{sec:direct-method} through
\ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

@@

Now we consider a Monte Carlo method that can be used to estimate the
solution to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First,
rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
component of $\vx$:
\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\ &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}
The terms in the Neumann series can be interpreted as a series of
transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk
with $k$ events that initiates in state $i$:
\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\ &=
    \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}
Here, the weight on the $m^\text{th}$ step is denoted $W_m$, and each
random walk starts with unit weight. At every step, contributions to
each component of $X$ are generated only in the starting state of the
history, $i_0$, by the source term in the current state,
$b_{i_m}$. Therefore, for a given random walk permutation, each state
must be used as a starting value in order to calculate its component
of the expectation value (i.e. a state of size $N$ will have $N$
histories per random walk permutation). When there is a transition
between states, for example $i\rightarrow j$, the weight is multiplied
by the transition factor
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
  \label{eq:weight}
\end{equation}
where $p_{ij}$ denotes the probability of transitioning from state $i$
to $j$. Then, the expected value of $X$ is
\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\ &=
    \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}
where $\nu$ denotes a particular random walk permutation.  Therefore,
the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased estimator
of the components of $\vx$ provided $\rho(\vH) < 1$.

@@

%%---------------------------------------------------------------------------%%
\subsection{Monte Carlo Synthetic Acceleration}
\label{sec:iter-refin-monte}

The methods presented in \S~\ref{sec:background} are characterized by slow
convergence rates bound by the Central Limit Theorem.
Halton~\cite{halton_1962,halton_1994} proposed a staged residual scheme called
Sequential Monte Carlo to speed up the convergence of these methods.  A
variant of this scheme has been successfully applied to the 1D nonlinear
thermal radiation diffusion equation in Ref.~\cite{evans_2003}.

Concisely, the Sequential Monte Carlo method solves Eq.~(\ref{eq:Ax=b}) using
the adjoint solution technique described in \S~\ref{sec:adjoint-method}.  The
following iteration scheme is applied:
\begin{subequations}
  \begin{gather}
    \vr^{l} = \vb - \vA\vx^{l}\:, \\ \vA\delta\vx^{l+1} =
    \vr^{l}\:, \label{eq:residual_solve}\\ \vx^{l+1} = \vx^{l} +
    \delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
In this scheme, the Monte Carlo adjoint method is used to estimate the
solution to Eq.~(\ref{eq:residual_solve}), and the residual is iterated to
convergence.  This iteration sequence is closely related to iterative
refinement, the exception being that there is no update to $x^{l+1}$ between
residual iterations.

We propose a modification of this scheme that uses Monte Carlo as a
synthetic acceleration for the fixed-point iteration given by
Eq.~(\ref{eq:richardson}). We begin by subtracting Eq.~(\ref{eq:richardson})
from Eq.~(\ref{eq:point_iteration}):
\begin{equation}
  \delta \vx^{l+1} = (\vI - \vA)\delta \vx^{l}\:,
  \label{eq:synthetic_error}
\end{equation}
where
\begin{equation}
  \delta \vx^{l} = \vx - \vx^{l}
\end{equation}
is defined as the error at iteration $l$. We then subtract $(\vI-\vA)
\delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}), giving a linear
system to solve for the error:
\begin{equation}
  \begin{split}
    \vA \delta \vx^{l+1} &= (\vI - \vA)(\vx^{l+1} - \vx^{l}) \\ &=
    \vr^{l+1}\:.
  \end{split}
  \label{eq:synthetic_error2}
\end{equation}

The following scheme will then converge in one iteration:
\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\ \vA \delta \vx^{l+1} =
    \vr^{l+1}\;,
    \label{eq:synthetic-residual-solve}\\ 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
To accelerate the fixed-point method and avoid inverting $\vA$ directly in
Eq.~(\ref{eq:synthetic-residual-solve}), we can instead use the above ideas to
create a new iterative scheme that applies the Monte Carlo method to estimate
the solution error. With this in hand, the Fixed-Point Monte Carlo
Synthetic-Acceleration (MCSA) method can be defined as follows:
\begin{subequations}
  \label{eq:MCSA-iteration}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\  \vr^{l+1/2} = \vb -
    \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\     
     \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
     \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
\end{subequations}
The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that the
Monte Carlo solution only approximately inverts this operator.  Thus, we have
defined a scheme in which the initial estimate of $\vx$ in each iteration is
updated using a single fixed-point iteration\footnote{A subspace method could
  be used instead of a stationary method here.}.  The residual is calculated
and used as a source to estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint method. The
error is then used to calculate an updated iterate of the solution vector,
$\vx^{l+1}$.  The entire sequence is iterated to convergence based on the
following stopping criterion~\cite{kelley_1995}:
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

The MCSA method defined in Eq.~(\ref{eq:MCSA-iteration}) uses the adjoint
method to estimate the error in Eq.~(\ref{eq:MCSA-residual_solve}) instead of
the direct method outlined in \S~\ref{sec:direct-method}. To demonstrate the
effectiveness of the adjoint method over the direct method within the context
of MCSA, we choose the 2D time-dependent Poisson equation as a
simple model problem:
\begin{equation}
  \frac{\partial \vu}{\partial t} = \nabla^2 \vu\:.
  \label{eq:poisson_equation}
\end{equation}
For all comparisons, a single time step is computed with backwards Euler time
integration. The Laplacian is differenced on a square Cartesian grid with a
second-order five-point stencil,
\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]\:,
  \label{eq:five_point_stencil}
\end{equation}
and a fourth-order nine-point stencil,
\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
  \label{eq:nine_point_stencil}
\end{multline}
both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
a single time step solution, we then have the following sparse linear system
to be solved with the MCSA method:
\begin{equation}
  \vA \vu^{n+1} = \vu^n\:.
  \label{eq:poisson_eq_lin_sys}
\end{equation}
Both the stencils will be used to vary the size and density of the sparse
linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).

A timing and convergence study is used to demonstrate the effectiveness of the
adjoint method as compared to the direct method. To assess both the CPU time
and number of iterations required to converge to a solution, a problem of
constant $\Delta$ was used with varying values of grid size, fixing the
spectral radius of the system at a constant value for each variation. Both the
five-point and nine-point stencils were used with both the direct and adjoint
solvers. For each case, 50 random walk permutations were computed per MCSA
iteration with a weight cutoff of \sn{1}{-4} for the stochastic linear solver
and a convergence tolerance of \sn{1}{-8} for the MCSA iterative solver. All
computations presented in this section were completed on a 2.66 GHz Intel Core
i7 machine with 8 GB 1067 MHz DDR3 memory. Figure~\ref{fig:poisson_cpu_time}
gives the CPU time needed for each case to converge in seconds, and
Fig.~\ref{fig:poisson_iterations} gives the number of iterations needed for
each case to converge relative to the problem size.
\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an $N \times
    N$ square mesh). Both the adjoint and direct solvers are used with the
    five point and nine point stencils. A significant CPU time speedup is
    noted with the adjoint method due to the reduced number of random walk
    events.}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.pdf}
  \caption{\sl Iterations to converge vs. Problem Size ($N$ for an $N \times
    N$ square mesh). Both the adjoint and direct solvers are used with the
    five-point and nine-point stencils. Both methods give similar iteration
    behavior.}
  \label{fig:poisson_iterations}
\end{figure}

We see clearly in Fig.~\ref{fig:poisson_cpu_time} that the using the adjoint
solver with MCSA results in several orders of magnitude speedup over the
direct solver while the number of iterations required to converge is of a
similar scale. We expect this for several reasons. First, with an equivalent
number of histories specified for both solvers and a system of size $N \times
N$, the direct solver will compute $N \times N$ random walks per permutation
while the adjoint solver will only compute one. This is necessary in the
direct method to ensure a contribution from each state as the random walk
sequence will only contribute to the starting state. For the adjoint method,
because the random walk sequence contributes to the state in which it
currently resides, fewer histories are neccessary to compute contributions
from all the nonzero parts of the source. Initially, this may not seem like a
fair comparison; however, we see from Fig.~\ref{fig:poisson_iterations} that
the number of iterations required to converge is approximately the same and
therefore the extra computations performed by the direct method do not improve
the error estimation. Therefore, although the total number of random walks per
iteration is in fact larger for the direct method, the solution computed at
each iteration is no better than that computed by the adjoint method.

As an additional comparison, the convergence behavior of MCSA can be analyzed
using both solvers to detect any performance benefits. To assess the
convergence properties of MCSA using each solver and stencil, the infinity
norm of the residual computed in Eq.~(\ref{eq:MCSA-residual}) was collected at
each iteration for a fixed problem size. Figure~\ref{fig:poisson_convergence}
gives the results of these computations. First, it is worthy to note on the
semilog plot that we are indeed achieving the expected exponential convergence
from MCSA. Second, we note that both the direct and adjoint methods have
approximately equivalent convergence behavior, leading to CPU time being the
dominating factor in method selection.
\begin{figure}[h!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.pdf}
  \caption{\sl Infinity norm of the solution residual vs. iteration number for
    a problem of fixed size. Both the adjoint and direct solvers are used with
    the five point and nine point stencils.}
  \label{fig:poisson_convergence}
\end{figure}
